---
title: "R Notebook"
output: html_notebook
---

Install and then import packages

```{r Install}
#source("https://bioconductor.org/biocLite.R")
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("biomaRt")
BiocManager::install("org.Hs.eg.db")
BiocManager::install("consensusOV")
```

Imports
```{r Imports}
library(biomaRt)
library(org.Hs.eg.db)
library(consensusOV)
library(dplyr)
library(genefu)
```


```{r Identify Files}
getwd()
subset_dir <- "data/gdc_download_20250330_115711.237016"

# Find all files ending in .tsv recursively
tsv_files <- list.files(
  path = subset_dir,
  pattern = "\\.tsv$",
  recursive = TRUE,
  full.names = TRUE
)

# Filter to make sure we're only keeping files (not directories or symlinks)
tsv_files <- tsv_files[file.info(tsv_files)$isdir == FALSE]

print(length(tsv_files))

```

```{r}
# Select only files with rna_seq in name
rna_files <- tsv_files[grepl("rna_seq", tsv_files)]
print(length(rna_files))
```


```{r Create Table}
# Loop through each file and extract the gene_id and tpm, which we can use to track all of the genes across all of the individuals

rna_list <- list()
for (file in rna_files) {
  temp_data <- read.delim(file, header = TRUE, sep = "\t", comment.char = "#")
  temp_data <- temp_data[, c("gene_id", "tpm_unstranded")] #gene_name, gene_type, raw read counts (for unstranded and stranded, maybe useful in normalization?) -- are less relevant in classification using consensusOV
  sample_name <- basename(file) 
  colnames(temp_data)[2] <- sample_name 
  rna_list[[sample_name]] <- temp_data
}
# This will create a table with gene_id as the first column and each sample
# in columns 2:
rna_expression_combined <- Reduce(function(x, y) merge(x, y, by = "gene_id", all = TRUE), rna_list)

# Create a column of cleaned ENSEMBL ids (remove version)
rna_expression_combined$ensembl_gene_id <- sub("\\.\\d+$", "",rna_expression_combined$gene_id)

```

```{r Map Ensembl IDs onto Entrez and Gene Symbol}
mart <- biomaRt::useMart(biomart = "ENSEMBL_MART_ENSEMBL",
                         dataset = "hsapiens_gene_ensembl",
                         host = "https://www.ensembl.org")

genes <- getBM(
  filters = "ensembl_gene_id",
  attributes = c("ensembl_gene_id", "entrezgene_id", "hgnc_symbol"),
  values = rna_expression_combined$ensembl_gene_id,
  mart = mart
)
```


```{r}
# Check for duplicated values in the entrez or gene symbol columns
genes %>%
  group_by(ensembl_gene_id) %>%
  summarise(
    n_entrez = n_distinct(entrezgene_id, na.rm = TRUE),
    n_symbol = n_distinct(hgnc_symbol, na.rm = TRUE),
    .groups = "drop"
    ) %>%
  filter(n_entrez > 1 | n_symbol > 1)

bad <- genes %>%
  group_by(entrezgene_id) %>%
  summarise(
    n_entrez = n_distinct(ensembl_gene_id, na.rm = TRUE),
    .groups = "drop"
    ) %>%
  filter(n_entrez > 1)

# Remove duplicated ENSEMBL --> entrez mappings by retaining the shortest
# entrez id (most likely to be a real gene rather than a LOC)
genes_clean <- genes %>%
  # Step 1: Keep shortest entrez ID per ensembl ID
  mutate(entrez_length = nchar(as.character(entrezgene_id))) %>%
  group_by(ensembl_gene_id) %>%
  slice_min(entrez_length, with_ties = FALSE) %>%
  ungroup() %>%
  select(-entrez_length) %>%

  # Step 2: Remove duplicate entrez IDs if the symbol is blank or NA
  group_by(entrezgene_id) %>%
  filter(!(n() > 1 & (hgnc_symbol == "" | is.na(hgnc_symbol)))) %>%
  ungroup() %>%

  # Step 3: For remaining entrezgene_id duplicates, keep the shortest symbol, or first alphabetically if tied
  mutate(symbol_length = nchar(hgnc_symbol)) %>%
  group_by(entrezgene_id) %>%
  arrange(symbol_length, hgnc_symbol) %>%
  slice(1) %>%
  ungroup() %>%
  select(-symbol_length)

  # Join to add 'name' from lookup table into rna_expression_combined
rna_expression_harmonized <- rna_expression_combined %>%
  left_join(genes_clean, by = "ensembl_gene_id")

# Remove rows where entrez_ids are NA or match the duplicated IDs to remove
rna_expression_filtered <- rna_expression_harmonized[!is.na(rna_expression_harmonized$entrezgene_id), ]

```

```{r}
# filtering the data. we only need geneid and expression for each patients in wide format
data_matrix <- as.matrix( rna_expression_filtered[, 2:425])

# Extract entrez_ids and assign to row names
rownames(data_matrix) <- rna_expression_filtered[, 427]

```


When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Cmd+Shift+K* to preview the HTML file). 
```{r}
bentink.subtypes <- get.subtypes(data_matrix, rownames(data_matrix), method = "Bentink")
bentink.subtypes$Bentink.subtypes
konecny.subtypes <- get.subtypes(data_matrix, rownames(data_matrix), method = "Konecny")
konecny.subtypes$Konecny.subtypes
helland.subtypes <- get.subtypes(data_matrix, rownames(data_matrix), method = "Helland")
helland.subtypes$Helland.subtypes


#conc.subtypes <- get.subtypes(data_matrix, rownames(data_matrix), "consensusOV")
```

```{r}
table(bentink.subtypes$Bentink.subtypes)
table(helland.subtypes$Helland.subtypes)
table(konecny.subtypes$Konecny.subtypes)
```

```{r}
# Assuming your matrix is called `consensus_matrix`
# and has the columns: IMR_consensus, DIF_consensus, PRO_consensus, MES_consensus

# Step 1: Get the max value per row
row_max <- apply(conc.subtypes$rf.probs, 1, max, na.rm = TRUE)

# Step 2: Plot histogram
hist(row_max,
     breaks = 20,
     col = "steelblue",
     main = "Histogram of Max Consensus Scores per Row",
     xlab = "Max Consensus Score",
     ylab = "Count")


```
```{r}
# Let's say your matrix is called `consensus_matrix`

consensus_matrix <- conc.subtypes$rf.probs

# Let's say your matrix is called `consensus_matrix`

# Step 1: Find the name of the column with the max value per row
max_col_names <- apply(consensus_matrix, 1, function(row) {
  colnames(consensus_matrix)[which.max(row)]
})

# Step 2: Create a 2-column dataframe with row names preserved
max_label_df <- data.frame(
  sample = rownames(consensus_matrix),
  max_label = max_col_names,
  row.names = NULL,  # optional: removes row numbers
  stringsAsFactors = FALSE
)


```

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.
```{r}

```

