---
title: "miRNA Preprocessing"
subtitle: "TCGA miRNA — load, filter, normalize (Hildana, June 2026)"
format: html
editor: visual
---

## Goal

Load raw TCGA miRNA read counts (497 samples), filter out non-informative miRNAs, and produce a normalized, log-transformed matrix for downstream clustering. The output `mirna_data/mirna_lcpm_matrix.rds` is the hand-off to `02_mirna_explore_clustering.qmd` and `03_mirna_nmf_final.qmd`.

```{r}
#| label: setup
#| message: false
library(tidyverse)
library(edgeR)

```

## Load & combine miRNA counts

```{r}
#| label: load-counts
mirna_dir   <- "TCGA-data_10:10:24_RNAandmiRNA"   # raw TCGA download (mRNA + miRNA)
mirna_files <- list.files(mirna_dir, pattern = "mirna.*\\.txt$",
                          recursive = TRUE, full.names = TRUE)

# Read one file -> two columns: miRNA_ID and read_count (renamed to the sample)
read_one_mirna <- function(file) {
  d <- read.delim(file, sep = "\t", comment.char = "#")
  if (!all(c("miRNA_ID", "read_count") %in% colnames(d))) {
    cat("Skipping file, missing columns:", file, "\n")
    return(NULL)
  }
  setNames(d[, c("miRNA_ID", "read_count")], c("miRNA_ID", basename(file)))
}

mirna_counts <- Reduce(function(x, y) merge(x, y, by = "miRNA_ID", all = TRUE),
                       Filter(Negate(is.null), lapply(mirna_files, read_one_mirna)))
mirna_counts[is.na(mirna_counts)] <- 0   # missing miRNA in a sample -> 0 reads

dim(mirna_counts)   # 1881 miRNA x (1 ID + 497 samples)
```

## Filter zero-sum miRNAs

Drop miRNAs with zero counts across **all** samples — they carry no information.

```{r}
#| label: filter-zeros
mirna_keep <- rowSums(mirna_counts[-1]) >= 1
cat("Dropped:", sum(!mirna_keep), "| Kept:", sum(mirna_keep), "\n")

mirna_counts <- mirna_counts[mirna_keep, ]

# Single transpose -> samples (rows) x miRNA (cols)
mirna_mat <- t(as.matrix(mirna_counts[-1]))
colnames(mirna_mat) <- mirna_counts$miRNA_ID
dim(mirna_mat)   # 497 x 1560
```

```{r}
#| label: qc-density
# Distribution of per-miRNA expression sums (log scale)
ggsave(
  "mirna_plots/mirna_expression_sum_density.png",
  ggplot(data.frame(s = colSums(mirna_mat)), aes(log(s))) +
    geom_density(fill = "pink", color = "black", alpha = 0.6) +
    labs(x = "log(Expression_Sum)", y = "Density") +
    theme_minimal(),
  width = 8, height = 6, dpi = 300
)
```

## Normalize & log-transform

`calcNormFactors` corrects for sequencing-depth differences between samples; `cpm(log = TRUE)` returns log2 counts-per-million.

```{r}
#| label: normalize
# QC: total counts per sample before normalization
mirna_sample_totals <- rowSums(mirna_mat)
summary(mirna_sample_totals)

png("mirna_plots/mirna_sample_total_counts_histogram.png", width = 1000, height = 800, res = 120)
hist(mirna_sample_totals, breaks = 50,
     main = "Distribution of Total miRNA Counts",
     xlab = "Total mirna Counts per Sample")
dev.off()

# edgeR expects genes-as-rows, so feed the miRNA x sample orientation
mirna_dge  <- calcNormFactors(DGEList(counts = t(mirna_mat)))
mirna_lcpm <- cpm(mirna_dge, log = TRUE, prior.count = 1)   # miRNA x sample
dim(mirna_lcpm)
```

## Save hand-off matrix

```{r}
#| label: save
saveRDS(mirna_lcpm, "mirna_data/mirna_lcpm_matrix.rds")   # miRNA x sample, normalized + log
```