---
title: "miRNA PCA"
subtitle: "TCGA miRNA — principal component analysis (Hildana, June 2026)"
format: html
editor: visual
---

## Goal

Explore structure in the normalized miRNA matrix with PCA. Diagnostic only — produces no cluster assignments. The scree plot shows how much variance each component captures, and the pairwise plot checks the first 5 components for separable structure.

```{r}
#| label: setup
#| message: false
library(tidyverse)
library(factoextra)
library(GGally)

```

## Load hand-off matrix

```{r}
#| label: load
mirna_lcpm <- readRDS("mirna_data/mirna_lcpm_matrix.rds")   # miRNA x sample, normalized + log
dim(mirna_lcpm)
```

## Run PCA

`prcomp` expects observations as rows, so we transpose to samples x miRNA.

```{r}
#| label: pca
mirna_pca <- prcomp(t(mirna_lcpm), center = TRUE, scale. = TRUE)
```

## Scree plot

```{r}
#| label: pca-scree
ggsave(
  "mirna_plots/mirna_pca_scree_plot.png",
  fviz_eig(mirna_pca, addlabels = TRUE, ylim = c(0, 100), main = ""),
  width = 8, height = 6, dpi = 300
)
```

## Pairwise plot of PC1–PC5

```{r}
#| label: pca-pairwise
mirna_pca_pairs <- as.data.frame(mirna_pca$x[, 1:5])
colnames(mirna_pca_pairs) <- paste0("PC", 1:5)

ggsave(
  "mirna_plots/mirna_pca_pairwise_pc1_to_pc5.png",
  ggpairs(mirna_pca_pairs, title = "Pairwise Plots of Principal Components 1-5"),
  width = 12, height = 12, dpi = 300
)
```

## Save PCA for reuse

The k-means file projects its clusters onto these PCA axes, so save the result rather than recomputing.

```{r}
#| label: save
saveRDS(mirna_pca, "mirna_data/mirna_pca.rds")
```