---
title: "miRNA subtype survival — multivariable Cox"
format: html
---

Does miRNA NMF subtype carry prognostic information **after** adjusting for
the clinical variables we already know matter (stage, grade, age)? The
log-rank test on the raw KM curves ignores those covariates; a multivariable
Cox model asks whether cluster membership adds anything on top of them.

The headline number is the **likelihood-ratio test** comparing a clinical-only
model to one that also includes the subtype — one p-value for "does subtype
add prognostic value?" — not the individual per-cluster p-values.

## Setup

```{r}
#| message: false
#| warning: false
library(readr)
library(dplyr)
library(stringr)
library(survival)
library(survminer)
library(ggplot2)

poster_cluster_palette <- c(
  "1" = "#4F6FA8",
  "2" = "#7F9CCB",
  "3" = "#7A7F87",
  "4" = "#D1837D",
  "5" = "#A83F4D"
)
poster_theme <- theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold"),
    legend.position = "right"
  )
```

## Load data

Cluster labels come from the harmonized `labels` table (`miRNA_nmf_cluster`);
survival + clinical covariates come from the sample-level subtype/clinical
table. Point `labels_file` at the k5 folder to run the same analysis for k=5.

```{r}
labels_file <- "data/gdc_tcga_ov_omics/harmonized_labels_k5/omics_sample_assay_availability_labels.tsv"

labels <- read_tsv(labels_file, show_col_types = FALSE)
clin   <- read.csv("mRNA_clusters/output/rna_subtypes_clinical_sample_level.csv")
```

## Build the per-case analysis table

Both inputs are sample-level, so we collapse to one row per case. Overall
survival time is `days_to_death` for deceased patients and
`days_to_last_follow_up` for living patients.

`residual_disease` — normally a key HGSOC prognostic covariate — is
**deliberately excluded**: in this GDC cart it is ~96% missing (only ~14 of
414 cases have a value), so including it collapses the complete-case model to
a dozen patients and it fails to converge. Stage and grade are collapsed
because the cohort is overwhelmingly advanced / high-grade (little spread).

```{r}
# one miRNA cluster per case
mi <- labels |>
  filter(has_miRNA, !is.na(miRNA_nmf_cluster)) |>
  group_by(case_submitter_id) |> slice(1) |> ungroup() |>
  transmute(case_submitter_id, mirna_cluster = factor(miRNA_nmf_cluster))

# survival + clinical covariates per case
df_all <- clin |>
  filter(!is.na(case_submitter_id)) |>
  group_by(case_submitter_id) |> slice(1) |> ungroup() |>
  mutate(
    event = if_else(vital_status == "Dead", 1L, 0L),
    time  = if_else(vital_status == "Dead",
                    as.numeric(days_to_death),
                    as.numeric(days_to_last_follow_up)),
    age   = as.numeric(age_at_index),
    stage_major = case_when(
      str_detect(figo_stage, "IV")  ~ "IV",
      str_detect(figo_stage, "III") ~ "III",
      str_detect(figo_stage, "II")  ~ "II",
      str_detect(figo_stage, "I")   ~ "I"),
    stage = case_when(stage_major %in% c("I", "II") ~ "I-II",
                      stage_major == "III"          ~ "III",
                      stage_major == "IV"           ~ "IV"),
    grade = case_when(tumor_grade == "G3"            ~ "G3",
                      tumor_grade %in% c("G1", "G2") ~ "G1-2",
                      TRUE                           ~ NA_character_)
  ) |>
  inner_join(mi, by = "case_submitter_id") |>
  mutate(
    mirna_cluster = relevel(factor(mirna_cluster), ref = "1"),  # reference cluster
    stage = factor(stage, levels = c("I-II", "III", "IV")),     # unordered
    grade = factor(grade, levels = c("G1-2", "G3"))             # unordered
  ) |>
  select(case_submitter_id, time, event, age, stage, grade, mirna_cluster) |>
  filter(!is.na(time), time >= 0, !is.na(event))

# KM only needs survival time, event status, and miRNA cluster.
df_km <- df_all |>
  select(case_submitter_id, time, event, mirna_cluster)

# Cox needs complete clinical covariates.
df <- df_all |>
  filter(!is.na(age), !is.na(stage), !is.na(grade))

cat("KM analysis n:", nrow(df_km), " events:", sum(df_km$event), "\n")
cat("Cox analysis n:", nrow(df), " events:", sum(df$event), "\n")
table(df$mirna_cluster); table(df$stage); table(df$grade)
```

## Kaplan-Meier curve

```{r}
#| fig-width: 8
#| fig-height: 6
km_fit <- survfit(Surv(time, event) ~ mirna_cluster, data = df_km)
km_palette <- unname(poster_cluster_palette[levels(df_km$mirna_cluster)])

km_plot <- ggsurvplot(
  km_fit,
  data = df_km,
  palette = km_palette,
  pval = TRUE,
  risk.table = TRUE,
  conf.int = FALSE,
  censor.shape = "|",
  censor.size = 2.5,
  title = "Overall survival by miRNA NMF cluster (k5)",
  xlab = "Days",
  ylab = "Overall survival probability",
  legend.title = "NMF cluster",
  legend.labs = paste0("C", levels(df_km$mirna_cluster)),
  ggtheme = poster_theme,
  tables.theme = theme_minimal(base_size = 10)
)

png("miRNA_clusters/mirna_plots/mirna_km_survival_k5_poster.png", width = 1000, height = 850, res = 120)
print(km_plot)
dev.off()

km_plot
```

## Nested Cox models + likelihood-ratio test

```{r}
fit_clin <- coxph(Surv(time, event) ~ stage + grade + age, data = df)
fit_full <- coxph(Surv(time, event) ~ stage + grade + age + mirna_cluster, data = df)

summary(fit_full)                 # per-term HR, 95% CI, p
anova(fit_clin, fit_full)         # HEADLINE: does cluster add prognostic value?
```

## Proportional-hazards check

```{r}
cox.zph(fit_full)
```

Note that a PH violation in `age`/`grade` does not invalidate the miRNA
conclusion: what matters is whether `mirna_cluster` violates PH and whether it
is significant. If you need clean adjusted HRs for the clinical effects,
stratify on the violating term (e.g. `+ strata(grade)`).

## Forest plot

Built by hand from the model's HR table — `survminer::ggforest()` is fragile
across versions (throws "names do not match previous names" here), so this is
the reliable equivalent.

```{r}
#| fig-width: 8
#| fig-height: 5
ci <- summary(fit_full)$conf.int
cf <- summary(fit_full)$coefficients
forest_tab <- data.frame(
  term = rownames(ci),
  HR   = ci[, "exp(coef)"],
  lo   = ci[, "lower .95"],
  hi   = ci[, "upper .95"],
  p    = cf[, "Pr(>|z|)"]
)

ggplot(forest_tab, aes(x = HR, y = reorder(term, HR))) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "grey50") +
  geom_linerange(aes(xmin = lo, xmax = hi)) +
  geom_point(size = 2) +
  scale_x_log10() +
  labs(x = "Hazard ratio (95% CI, log scale)", y = NULL,
       title = "Multivariable Cox: miRNA subtype + stage/grade/age") +
  theme_minimal()
```

## Interpretation

- **LRT significant + `mirna_cluster` PH holds** → miRNA subtype is an
  independent prognostic factor after adjustment.
- **LRT not significant** → the null is robust; subtype adds nothing beyond
  stage/grade/age. Report honestly alongside the KM — a legitimate result
  consistent with HGSOC's clinically driven outcomes.
- **`mirna_cluster` PH violated** → don't report a single HR for it; stratify,
  add a time interaction, or use a time-restricted comparison.
