---
title: "preprocessing_and_nmf_cluster"
format: html
editor: visual
---

```{r}
#install.packages(c("tidyverse", "DESeq2", "umap"))
library(tidyverse)

#if (!requireNamespace("BiocManager", quietly = TRUE)) {
#  install.packages("BiocManager")
#}
#BiocManager::install("Biobase") 
#BiocManager::install("NMF")

library(Biobase)
library(umap)
#library(consensusOV)
#library(TCGAbiolinks)
library(dplyr)
library(reshape2)
library(ggplot2)
library(factoextra)
library(NMF)
library(GGally)
```

```{r}
full_dir <- "/home/hms68/project_pi_rf273/hms68/miRNA_OC/miRNA_jan_26/TCGA-data_10:10:24_RNAandmiRNA"
txt_files <- list.files(full_dir, pattern = "\\.txt$", recursive = TRUE, full.names = TRUE)
mirna_files <- txt_files[grepl("mirna", txt_files)]
```

```{r}
#load miRNA data
mirna_list <- list()
for (file in mirna_files) {
  temp_data <- read.delim(file, header = TRUE, sep = "\t", comment.char = "#") # confirm that we have the column we need 
  if (!all(c("miRNA_ID", "read_count") %in% colnames(temp_data))) {
    cat("Skipping file, missing columns:", file, "\n")
    next
  }
  
  temp_data <- temp_data[, c("miRNA_ID", "read_count")] #select the columns we need
  
  sample_name <- basename(file) #rename column
  colnames(temp_data)[2] <- sample_name
  mirna_list[[sample_name]] <- temp_data #store data
}
mirna_expression_combined <- Reduce(function(x, y) merge(x, y, by = "miRNA_ID", all = TRUE), mirna_list) # merge to combined expression matrix
mirna_expression_combined[is.na(mirna_expression_combined)] <- 0 # any NA to 0, representing missing value
```

Transposing the data (patients as rows and miRNAs as columns)

```{r}
#write.csv(mirna_expression_combined, row.names = FALSE)
miRNA_ids <- mirna_expression_combined$miRNA_ID
miRNA_data_numeric <- mirna_expression_combined[, -1]
miRNA_data_t <- as.data.frame(t(miRNA_data_numeric))
colnames(miRNA_data_t) <- miRNA_ids
rownames(miRNA_data_t) <- rownames(t(miRNA_data_numeric))
head(miRNA_data_t)
```

sum

```{r}
miRNA_sums <- apply(miRNA_data_t, 2, sum, na.rm = TRUE)
miRNA_sums_df <- data.frame(miRNA_ID = names(miRNA_sums), Expression_Sum = miRNA_sums)
miRNA_sums_zero <- miRNA_sums_df |>
  filter(Expression_Sum < 1) #321 miRNAs are zero, this makes sense because they all have 1881 miRNA (there are nearly 2,800 human miRNAs but I am not sure why we are only using 1881) 

#filter out non-zero miRNA sums
miRNA_sums_nonzero <- miRNA_sums_df |>
  filter(Expression_Sum >= 1) #1560 miRNAs are non-zero Mean square error, SD, RMSD, metrics for error 
```

```{r}
library(ggplot2)
#visualize with density plot of the log-transformed sum 
ggplot(miRNA_sums_nonzero, aes(x = log(Expression_Sum))) +
  geom_density(fill = "pink", color = "black", alpha = 0.6) +
  labs(x = "log(Expression_Sum)", y = "Density") + theme_minimal()
```

```{r}
#filter miRNAs identified as non-zero in the original data
nonzero_miRNAs <- miRNA_sums_nonzero$miRNA_ID

filtered_mirna_expression_combined <- mirna_expression_combined %>%
  filter(miRNA_ID %in% nonzero_miRNAs)

for (patient in colnames(filtered_mirna_expression_combined)[-1]) { # exclude the miRNA_ID column
  output_file <- paste0(full_dir, "/filtered_", patient, ".txt")
  write.table(filtered_mirna_expression_combined[, c("miRNA_ID", patient)], file = output_file, sep = "\t", row.names = FALSE, quote = FALSE)
}
```

```{r}
#miRNA_data_t_com_nor is Transposing combined data and converting to df 
miRNA_ids_com <- filtered_mirna_expression_combined$miRNA_ID
miRNA_data_numeric_com <- filtered_mirna_expression_combined[, -1]
#miRNA_data_t_com_nor <- as.data.frame(t(miRNA_data_numeric_com))
miRNA_data_t_com_nor <- as.data.frame((miRNA_data_numeric_com))
colnames(miRNA_data_t_com_nor) <- miRNA_ids_com
rownames(miRNA_data_t_com_nor) <- rownames(t(miRNA_data_numeric_com))
```

PCA - data explorarion

```{r}
#dataframe 
miRNA_data_t_com_nor_df <- as.data.frame(miRNA_data_t_com_nor)

perform_pca <- function(data, scale_data = TRUE) {
  pca_result <- prcomp(data, center = TRUE, scale. = scale_data)
  return(pca_result)
}

miRNA_data_pca4 <- perform_pca(miRNA_data_t_com_nor_df)
fviz_eig(miRNA_data_pca4, addlabels = TRUE, ylim = c(0, 50),
         main = "") 
```

```{r}
#create a data frame of the first 5 principal components
pca_pairwise_data <- as.data.frame(miRNA_data_pca4$x[, 1:5])
colnames(pca_pairwise_data) <- paste0("PC", 1:5)

#pairwise plot of the first 5 principal components
ggpairs(pca_pairwise_data,
        title = "Pairwise Plots of Principal Components 1-5")
```

```{r}
explained_variance <- cumsum(miRNA_data_pca4$sdev^2 / sum(miRNA_data_pca4$sdev^2)) * 100 #80% explained with close to 200 PCs
```

Clustering - (using k-means)

```{r}
set.seed(123)
number_of_clusters <- 4  #assume there are four clusters, what would they look like using miRNA?
kmeans_result <- kmeans(miRNA_data_t_com_nor, centers = number_of_clusters, nstart = 25)
#kmeans_result <- kmeans(filtered_miRNA_data_by_hybrid, centers = number_of_clusters, nstart = 25)

#visualize the clusters
pca_data <- as.data.frame(miRNA_data_pca4$x) 
pca_data$Cluster <- as.factor(kmeans_result$cluster)

ggplot(pca_data, aes(x = PC1, y = PC2, color = Cluster)) + geom_point(size = 3, alpha = 0.7) +
  labs( x = "Principal Component 1", y = "Principal Component 2") + theme_minimal() + theme(legend.position = "right")
```

potential another approach for filtering - using hybrid mean and **high mean and high variability first**

```{r}

```

Clustering - using NMF

```{r}
set.seed(42) 
k_range <- 2:8 #assuming the clusters will be in the rangg 2 - 8
#nmf_result <- nmf(miRNA_data_t_com_nor, rank = k_range, method = "brunet", nrun = 50, .opt = "v-p")
nmf_result <- nmf(miRNA_data_t_com_nor, rank = 4, method = "brunet", nrun = 1, .opt = "v-p")

#plot to visualuze 
consensusmap(nmf_result, main = "Consensus NMF Clustering")a

```

We can also use other clustering methods - including but not limited to Gaussian Mixture Models (GMM) which unlike k-means uses soft assignments.

```{r}
#get the assigned cluster members 
#get cluster assignments for best rank
#best_model <- fit(nmf_result, rank = 4)  
best_model <- fit(nmf_result) 
clusters <- apply(coef(best_model), 2, which.max)

table(clusters) #view assignments
split(names(clusters), clusters) #get members of each cluster
```
