---
title: "mRNA subtype survival (Kaplan–Meier)"
format: html
---

Kaplan–Meier overall-survival curves for each mRNA subtyping method
(consensusOV, Konecny, Helland, Verhaak, Bentink). All subtype calls and
the survival fields live in the same sample-level table, so no joins are
needed — we just collapse to one row per case and derive overall survival.

## Setup

```{r}
#| message: false
#| warning: false
library(readr)
library(dplyr)
library(survival)
library(survminer)
```

## Load data

```{r}
mrna_clinical_cluster <- read.csv(
  "mRNA_clusters/output/rna_subtypes_clinical_sample_level.csv"
)
```

## Case-level survival

The table is sample-level (a case can have more than one RNA aliquot), so we
keep one row per case, preferring the primary-tumor sample (`sample_type_code`
`01`). Overall survival time is `days_to_death` for deceased patients and
`days_to_last_follow_up` (last-contact time) for living patients.

```{r}
methods <- c(
  consensusOV = "consensusOV_subtype",
  konecny     = "konecny_subtype",
  helland     = "helland_subtype",
  verhaak     = "verhaak_subtype",
  bentink     = "bentink_subtype"
)

surv_df <- mrna_clinical_cluster |>
  filter(!is.na(case_submitter_id)) |>
  # prefer primary tumor (sample_type_code may load as "01" or 1)
  mutate(is_primary = suppressWarnings(as.integer(sample_type_code)) == 1L) |>
  arrange(case_submitter_id, desc(is_primary)) |>
  group_by(case_submitter_id) |>
  slice(1) |>
  ungroup() |>
  mutate(
    os_event = if_else(vital_status == "Dead", 1L, 0L),
    os_time  = if_else(vital_status == "Dead",
                       as.numeric(days_to_death),
                       as.numeric(days_to_last_follow_up))
  ) |>
  filter(!is.na(os_time), !is.na(os_event))

cat("Cases with usable survival:", nrow(surv_df), "\n")
cat("Events (deaths):", sum(surv_df$os_event), "\n")
```

## KM curves per subtyping method

```{r}
#| fig-width: 8
#| fig-height: 7
for (m in names(methods)) {
  col <- methods[[m]]
  d <- surv_df |> filter(!is.na(.data[[col]]))

  # surv_fit() (survminer) stores the formula + data so ggsurvplot works
  # correctly inside a loop.
  fit <- surv_fit(
    as.formula(paste0("Surv(os_time, os_event) ~ ", col)),
    data = d
  )

  p <- ggsurvplot(
    fit, data = d,
    pval = TRUE, risk.table = TRUE, conf.int = FALSE,
    title = paste0("mRNA subtype: ", m, " (n = ", nrow(d), ")"),
    xlab = "Days", ylab = "Overall survival",
    legend.title = m
  )
  print(p)
}
```

## Log-rank summary

```{r}
logrank_p <- vapply(methods, function(col) {
  d <- surv_df |> filter(!is.na(.data[[col]]))
  sd <- survdiff(
    as.formula(paste0("Surv(os_time, os_event) ~ ", col)),
    data = d
  )
  pchisq(sd$chisq, length(sd$n) - 1, lower.tail = FALSE)
}, numeric(1))

tibble(
  method    = names(methods),
  n_groups  = vapply(methods, function(col) dplyr::n_distinct(surv_df[[col]], na.rm = TRUE), integer(1)),
  logrank_p = signif(logrank_p, 3)
) |>
  arrange(logrank_p)
```
