---
title: "miRNA vs mRNA subtype concordance (ARI)"
format: html
---

Do the miRNA NMF clusters recapitulate the established mRNA subtypes? And does
k=5 agree any better than k=4? This compares each miRNA clustering against all
five mRNA subtyping methods using the **Adjusted Rand Index (ARI)**.

ARI is the right metric here: it handles different cluster counts (k=4 vs k=5)
and arbitrary integer labels, and it is **chance-corrected** (0 = agreement no
better than random, 1 = identical partitions). Cramér's V is reported
alongside, but note it is *not* chance-corrected — with n≈414 it can show a
small "significant" association even when ARI says there is essentially no
cluster-to-subtype mapping. **ARI is the headline.**

## Setup

```{r}
#| message: false
#| warning: false
library(readr)
library(dplyr)
library(ggplot2)
```

## Concordance metrics

```{r}
# Adjusted Rand Index between two labelings of the same cases.
ari <- function(a, b) {
  tab <- table(a, b); n <- sum(tab)
  sc  <- function(x) sum(choose(x, 2))
  idx <- sum(choose(tab, 2)); ai <- rowSums(tab); bj <- colSums(tab)
  expected <- sc(ai) * sc(bj) / choose(n, 2)
  maxidx   <- 0.5 * (sc(ai) + sc(bj))
  (idx - expected) / (maxidx - expected)
}

# Cramér's V (effect size for the contingency table); not chance-corrected.
cramersV <- function(a, b) {
  tab <- table(a, b)
  chi <- suppressWarnings(chisq.test(tab)$statistic)
  n <- sum(tab); k <- min(nrow(tab), ncol(tab))
  sqrt(as.numeric(chi) / (n * (k - 1)))
}
```

## Load data

mRNA subtype calls (5 methods) come from the sample-level table; miRNA cluster
labels come from the harmonized `labels` tables (k=4 and k=5 folders).

```{r}
methods <- c(consensusOV = "consensusOV_subtype", konecny = "konecny_subtype",
             helland = "helland_subtype", verhaak = "verhaak_subtype",
             bentink = "bentink_subtype")

# one mRNA call per case (prefer primary tumor)
mrna <- read.csv("mRNA_clusters/output/rna_subtypes_clinical_sample_level.csv") |>
  filter(!is.na(case_submitter_id)) |>
  mutate(pri = ifelse(suppressWarnings(as.integer(sample_type_code)) == 1L, 0L, 1L)) |>
  arrange(case_submitter_id, pri) |>
  group_by(case_submitter_id) |> slice(1) |> ungroup() |>
  select(case_submitter_id, all_of(unname(methods)))

# one miRNA cluster per case, for a given labels file
mirna_k <- function(path) {
  read_tsv(path, show_col_types = FALSE) |>
    filter(has_miRNA, !is.na(miRNA_nmf_cluster)) |>
    group_by(case_submitter_id) |> slice(1) |> ungroup() |>
    transmute(case_submitter_id, cluster = as.character(miRNA_nmf_cluster))
}
mi4 <- mirna_k("data/gdc_tcga_ov_omics/harmonized_labels/omics_sample_assay_availability_labels.tsv")
mi5 <- mirna_k("data/gdc_tcga_ov_omics/harmonized_labels_k5/omics_sample_assay_availability_labels.tsv")
```

## Score every method against a miRNA clustering

```{r}
score <- function(mi) {
  do.call(rbind, lapply(names(methods), function(m) {
    col <- methods[[m]]
    d <- mrna |> select(case_submitter_id, sub = all_of(col)) |>
      inner_join(mi, by = "case_submitter_id") |>
      filter(!is.na(sub), sub != "")
    data.frame(method = m, n = nrow(d),
               ARI = round(ari(d$cluster, d$sub), 3),
               CramersV = round(cramersV(d$cluster, d$sub), 3))
  }))
}

k4 <- score(mi4)
k5 <- score(mi5)
```

## Side-by-side comparison (k=4 vs k=5)

```{r}
data.frame(
  method = k4$method,
  ARI_k4 = k4$ARI, ARI_k5 = k5$ARI,
  V_k4 = k4$CramersV, V_k5 = k5$CramersV
)
```

## Concordance heatmap (k=5)

Cross-tabulation of miRNA k=5 clusters (rows) against each mRNA method
(columns). Cells are colored by **row percentage** (what fraction of each
miRNA cluster falls in that mRNA subtype) so the big clusters don't dominate
the scale; the number in each cell is the raw patient count. Concordance would
show as a diagonal (each row lighting up one column) — its absence is the
result.

```{r}
#| fig-width: 23
#| fig-height: 4.5
heat <- do.call(rbind, lapply(names(methods), function(m) {
  col <- methods[[m]]
  d <- mrna |> select(case_submitter_id, sub = all_of(col)) |>
    inner_join(mi5, by = "case_submitter_id") |>
    filter(!is.na(sub), sub != "")
  a <- ari(d$cluster, d$sub)
  d |> count(cluster, sub, name = "n") |>
    group_by(cluster) |> mutate(pct = 100 * n / sum(n)) |> ungroup() |>
    mutate(method = sprintf("%s (ARI=%.3f)", m, a))
}))
heat$cluster <- factor(heat$cluster, levels = rev(sort(unique(heat$cluster))))

ggplot(heat, aes(sub, cluster, fill = pct)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = n), size = 3) +
  facet_wrap(~ method, scales = "free_x", nrow = 1) +
  scale_fill_gradient(low = "#F7F4F0", high = "#B94A48", name = "% of\nmiRNA\ncluster") +  # poster_mid -> poster_red
  labs(x = "mRNA subtype", y = "miRNA NMF cluster (k=5)",
       title = "miRNA (k=5) vs mRNA subtype concordance",
       subtitle = "cell = row % within miRNA cluster; number = patient count") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size = 8),
        panel.spacing.x = unit(4, "lines"))
```

## Interpretation

- **ARI ≈ 0 for every method, both k** → miRNA NMF clusters do not recapitulate
  the mRNA subtypes; overlap is no better than chance.
- **k=5 ≈ k=4** (differences ±0.01, i.e. noise) → the extra cluster buys no
  additional alignment with mRNA structure.
- **Verhaak is marginally highest** on both ARI and Cramér's V — consistent
  with it being the one mRNA method with a KM signal — but the value is still
  negligible.
- **Bentink is at/below chance** because it has only 2 groups
  (Angiogenic / non-Angiogenic), so ARI against a 4–5 cluster labeling is ~0.

Taken with the survival results (weak KM separation, no independent Cox
effect), the miRNA clustering appears to capture structure that is largely
orthogonal to the established mRNA taxonomy rather than reproducing it.
