---
title: "miRNA K-means Clustering"
subtitle: "TCGA miRNA — k-means cross-check (Hildana, June 2026)"
format: html
editor: visual
---

## Goal

Cluster the normalized miRNA matrix with k-means (k = 4) as a distance-based cross-check against NMF. Clusters are projected onto the PCA axes from `02_mirna_pca.qmd` for visualization. K-means shows no clear separation on PC1/PC2, which motivates using NMF as the final method.

```{r}
#| label: setup
#| message: false
library(tidyverse)

project_root <- normalizePath(
  if (basename(getwd()) == "miRNA_clusters") file.path(getwd(), "..") else getwd(),
  mustWork = TRUE
)
mirna_data_dir <- file.path(project_root, "data", "mirna_data")
mirna_plot_dir <- file.path(project_root, "miRNA_clusters", "mirna_plots")
dir.create(mirna_plot_dir, showWarnings = FALSE, recursive = TRUE)
```

## Load hand-off matrix & PCA

```{r}
#| label: load
mirna_lcpm <- readRDS(file.path(mirna_data_dir, "mirna_lcpm_matrix.rds"))   # miRNA x sample, normalized + log
mirna_pca  <- readRDS(file.path(mirna_data_dir, "mirna_pca.rds"))           # from 02_mirna_pca.qmd
mirna_sample_metadata <- readr::read_csv(
  file.path(mirna_data_dir, "mirna_sample_metadata.csv"),
  show_col_types = FALSE
)
```

## Run k-means

`kmeans` clusters the samples, so we transpose to samples x miRNA.

```{r}
#| label: kmeans
set.seed(123)
mirna_kmeans <- kmeans(t(mirna_lcpm), centers = 4, nstart = 25)

table(mirna_kmeans$cluster)   # cluster sizes
```

## Visualize on PCA axes

```{r}
#| label: kmeans-plot
mirna_kmeans_plot_df <- as.data.frame(mirna_pca$x)
mirna_kmeans_plot_df$cluster <- factor(mirna_kmeans$cluster)

ggsave(
  file.path(mirna_plot_dir, "mirna_kmeans_pca_clusters.png"),
  ggplot(mirna_kmeans_plot_df, aes(PC1, PC2, color = cluster)) +
    geom_point(size = 3, alpha = 0.7) +
    labs(x = "PC 1", y = "PC 2") +
    theme_minimal() + theme(legend.position = "right"),
  width = 8, height = 6, dpi = 300
)
```

## Save cluster assignments

```{r}
#| label: save
mirna_kmeans_df <- data.frame(
  file_name = rownames(t(mirna_lcpm)),
  cluster   = as.integer(mirna_kmeans$cluster)
) |>
  left_join(mirna_sample_metadata, by = "file_name") |>
  relocate(
    file_name,
    file_id,
    case_submitter_id,
    case_id,
    sample_submitter_id,
    sample_type,
    sample_type_code,
    aliquot_barcode,
    cluster
  )
write.csv(mirna_kmeans_df, file.path(mirna_data_dir, "mirna_kmeans_subtypes_k4.csv"), row.names = FALSE)
```

