---
title: "miRNA NMF Clustering"
subtitle: "TCGA miRNA — NMF rank survey + final k=4 subtypes (Hildana, June 2026)"
format: html
editor: visual
---

## Goal

Cluster the downloaded GDC miRNA samples into subtypes by NMF of their miRNA profiles. This file does the rank survey (2–8) that justifies the cluster count, then fits the final model at k = 4 and writes per-sample subtype assignments with GDC/clinical identifiers. The main **deliverable** file is `miRNA_clusters/mirna_nmf_subtypes_k4.csv`; a copy is also written to `data/mirna_data/mirna_nmf_subtypes_k4.csv`.


```{r}
#| label: setup
#| message: false
library(tidyverse)
library(NMF)
NMF::nmf.options(cores = 4)

project_root <- normalizePath(
  if (basename(getwd()) == "miRNA_clusters") file.path(getwd(), "..") else getwd(),
  mustWork = TRUE
)
mirna_data_dir <- file.path(project_root, "data", "mirna_data")
mirna_plot_dir <- file.path(project_root, "miRNA_clusters", "mirna_plots")
mirna_cluster_dir <- file.path(project_root, "miRNA_clusters")
dir.create(mirna_plot_dir, showWarnings = FALSE, recursive = TRUE)
```

## Load hand-off matrix

```{r}
#| label: load
mirna_lcpm <- readRDS(file.path(mirna_data_dir, "mirna_lcpm_matrix.rds"))   # miRNA x sample
mirna_sample_metadata <- readr::read_csv(
  file.path(mirna_data_dir, "mirna_sample_metadata.csv"),
  show_col_types = FALSE
)
dim(mirna_lcpm)
```

## Make non-negative

NMF requires non-negative input, but `log = TRUE` CPM can be negative. Shift the whole matrix up by its minimum.

```{r}
#| label: nonneg
mirna_nmf_input <- mirna_lcpm - min(mirna_lcpm)
range(mirna_nmf_input)   # confirm lower bound is now 0
```

## NMF rank survey (2–8)

Runs NMF across candidate ranks to inform the cluster count (cophenetic correlation, dispersion, silhouette). This is the **slowest** step — cached to `.rds` so you only run it once.

```{r}
#| label: rank-survey
mirna_rank_survey_path <- file.path(mirna_data_dir, "mirna_nmf_rank_survey.rds")

if (file.exists(mirna_rank_survey_path)) {
  mirna_nmf_estim <- readRDS(mirna_rank_survey_path)
} else {
  set.seed(123)
  mirna_nmf_estim <- nmf(mirna_nmf_input, rank = 2:8, nrun = 50,
                         method = "brunet", .opt = "v-p")
  saveRDS(mirna_nmf_estim, mirna_rank_survey_path)
}

png(file.path(mirna_plot_dir, "mirna_nmf_rank_survey.png"), width = 1000, height = 800, res = 120)
plot(mirna_nmf_estim)
dev.off()
```

## Final NMF (k = 4)

k = 4 was chosen from the rank survey above (see thesis) and mirrors the 4 consensusOV ovarian subtypes. To explore another k, change `rank` and re-run the chunks below.

```{r}
#| label: nmf-final
set.seed(123)
mirna_nmf_result <- nmf(mirna_nmf_input, rank = 4, method = "brunet",
                        nrun = 50, .opt = "vp4")
saveRDS(mirna_nmf_result, file.path(mirna_data_dir, "mirna_nmf_result_k4.rds"))

# NOTE: mirna_nmf_input is miRNA x sample, so predict() clusters the 497 samples.
# To reproduce the original per-miRNA result instead, use t(mirna_nmf_input) here.
```

## Consensus plot

```{r}
#| label: consensus-plot
mirna_cons      <- NMF::consensus(mirna_nmf_result)
mirna_clusters  <- predict(mirna_nmf_result)
mirna_ord       <- order(mirna_clusters)
mirna_cons_sort <- mirna_cons[mirna_ord, mirna_ord]

png(file.path(mirna_plot_dir, "mirna_consensus_nmf.png"), width = 1000, height = 900)
image(
  x = seq_len(ncol(mirna_cons_sort)),
  y = seq_len(nrow(mirna_cons_sort)),
  z = t(mirna_cons_sort[nrow(mirna_cons_sort):1, ]),
  axes = FALSE, xlab = "Samples", ylab = "Samples",
  main = "Consensus NMF Clustering"
)
dev.off()
```

## Save subtype assignments

```{r}
#| label: save
mirna_clusters <- predict(mirna_nmf_result)
mirna_nmf_df   <- data.frame(
  file_name = names(mirna_clusters),
  cluster   = as.integer(mirna_clusters)
) |>
  left_join(mirna_sample_metadata, by = "file_name") |>
  relocate(
    file_name,
    file_id,
    case_submitter_id,
    case_id,
    sample_submitter_id,
    sample_type,
    sample_type_code,
    aliquot_barcode,
    cluster
  )
write.csv(mirna_nmf_df, file.path(mirna_cluster_dir, "mirna_nmf_subtypes_k4.csv"), row.names = FALSE)
write.csv(mirna_nmf_df, file.path(mirna_data_dir, "mirna_nmf_subtypes_k4.csv"), row.names = FALSE)

nrow(mirna_nmf_df)   
```

## Sanity checks

```{r}
#| label: sanity
head(mirna_nmf_df)
table(mirna_nmf_df$cluster)          

mirna_sample_totals <- colSums(mirna_nmf_input)
png(file.path(mirna_plot_dir, "mirna_counts_by_cluster.png"), width = 1000, height = 800, res = 120)
boxplot(mirna_sample_totals ~ mirna_nmf_df$cluster,
        ylab = "total miRNA counts", xlab = "cluster")
dev.off()
```
