---
title: "miRNA NMF Clustering"
subtitle: "TCGA miRNA — NMF rank survey + final subtypes (Hildana, June 2026)"
format: html
editor: visual
---

## Goal

Cluster the downloaded GDC miRNA samples into subtypes by NMF of their miRNA profiles. This file does the rank survey (2–8) that justifies the cluster count, then fits the final model at `nmf_k` clusters (set in the setup chunk) and writes per-sample subtype assignments with GDC/clinical identifiers. The main **deliverable** file is `data/mirna_data/mirna_nmf_subtypes_k<nmf_k>.csv`.

```{r}
#| label: setup
#| message: false
library(tidyverse)
library(NMF)
NMF::nmf.options(cores = 4)

project_root <- normalizePath(
  if (basename(getwd()) == "miRNA_clusters") file.path(getwd(), "..") else getwd(),
  mustWork = TRUE
)
mirna_data_dir <- file.path(project_root, "data", "mirna_data")
mirna_plot_dir <- file.path(project_root, "miRNA_clusters", "mirna_plots")
dir.create(mirna_plot_dir, showWarnings = FALSE, recursive = TRUE)

# Cluster count for the final NMF fit. Set to 4 to reproduce the original
# k=4 subtypes, 5 for k=5, etc. All output filenames below are suffixed with
# this value (e.g. _k5), so different k never overwrite each other.
nmf_k <- 5
k_tag <- paste0("k", nmf_k)

# Poster color system shared across NMF consensus, marker heatmaps, rank
# survey, and survival curves.
poster_blue <- "#6E8DBF"
poster_mid <- "#F7F4F0"
poster_red <- "#B94A48"
poster_cluster_palette <- c(
  "1" = "#4F6FA8",
  "2" = "#7F9CCB",
  "3" = "#7A7F87",
  "4" = "#D1837D",
  "5" = "#A83F4D"
)
poster_measure_palette <- c(
  "Basis" = "#A83F4D",
  "Best fit" = "#4F6FA8",
  "Coefficients" = "#7A7F87",
  "Consensus" = "#D1837D"
)
```

## Load hand-off matrix

```{r}
#| label: load
mirna_lcpm <- readRDS(file.path(mirna_data_dir, "mirna_lcpm_matrix.rds"))   # miRNA x sample
mirna_sample_metadata_path <- file.path(mirna_data_dir, "mirna_sample_metadata.csv")
mirna_sample_metadata <- if (file.exists(mirna_sample_metadata_path)) {
  readr::read_csv(mirna_sample_metadata_path, show_col_types = FALSE)
} else {
  tibble(file_name = colnames(mirna_lcpm))
}
dim(mirna_lcpm)
```

## Make non-negative

NMF requires non-negative input, but `log = TRUE` CPM can be negative. Shift the whole matrix up by its minimum.

```{r}
#| label: nonneg
mirna_nmf_input <- mirna_lcpm - min(mirna_lcpm)
range(mirna_nmf_input)   # confirm lower bound is now 0
```

## NMF rank survey (2–8)

Runs NMF across candidate ranks to inform the cluster count (cophenetic correlation, dispersion, silhouette). This is the **slowest** step — cached to `.rds` so you only run it once.

```{r}
#| label: rank-survey
mirna_rank_survey_path <- file.path(mirna_data_dir, "mirna_nmf_rank_survey.rds")

if (file.exists(mirna_rank_survey_path)) {
  mirna_nmf_estim <- readRDS(mirna_rank_survey_path)
} else {
  set.seed(123)
  mirna_nmf_estim <- nmf(mirna_nmf_input, rank = 2:8, nrun = 50,
                         method = "brunet", .opt = "v-p")
  saveRDS(mirna_nmf_estim, mirna_rank_survey_path)
}

png(file.path(mirna_plot_dir, "mirna_nmf_rank_survey.png"), width = 1000, height = 800, res = 120)
mirna_rank_plot <- plot(mirna_nmf_estim)
if (inherits(mirna_rank_plot, "ggplot")) {
  mirna_rank_plot <- mirna_rank_plot +
    scale_color_manual(values = poster_measure_palette) +
    theme_minimal(base_size = 12) +
    theme(
      strip.background = element_rect(fill = "#E8E8E8", color = "grey35"),
      panel.grid.minor = element_blank(),
      plot.title = element_text(face = "bold")
    )
}
print(mirna_rank_plot)
dev.off()
```

## Final NMF (k = `r nmf_k`)

k = 4 was chosen from the rank survey above (see thesis) and mirrors the 4 consensusOV ovarian subtypes. To explore another k, change `nmf_k` in the setup chunk and re-run the chunks below.

```{r}
#| label: nmf-final
set.seed(123)
mirna_nmf_result <- nmf(mirna_nmf_input, rank = nmf_k, method = "brunet",
                        nrun = 50, .opt = "vp4")
saveRDS(mirna_nmf_result, file.path(mirna_data_dir, paste0("mirna_nmf_result_", k_tag, ".rds")))

# NOTE: mirna_nmf_input is miRNA x sample, so predict() clusters the 497 samples.
# To reproduce the original per-miRNA result instead, use t(mirna_nmf_input) here.
```

## Consensus plot

```{r}
#| label: consensus-plot
mirna_cons      <- NMF::consensus(mirna_nmf_result)
mirna_clusters  <- predict(mirna_nmf_result)
mirna_ord       <- order(mirna_clusters)
mirna_cons_sort <- mirna_cons[mirna_ord, mirna_ord]
mirna_cluster_sizes <- table(mirna_clusters[mirna_ord])
mirna_cluster_boundaries <- cumsum(as.numeric(mirna_cluster_sizes)) + 0.5
mirna_cluster_midpoints <- cumsum(as.numeric(mirna_cluster_sizes)) -
  as.numeric(mirna_cluster_sizes) / 2 + 0.5
consensus_palette <- colorRampPalette(c(poster_blue, poster_mid, poster_red))(101)

png(file.path(mirna_plot_dir, paste0("mirna_consensus_nmf_", k_tag, ".png")), width = 1000, height = 900)
op <- par(mar = c(4, 4, 3, 1))
image(
  x = seq_len(ncol(mirna_cons_sort)),
  y = seq_len(nrow(mirna_cons_sort)),
  z = t(mirna_cons_sort[nrow(mirna_cons_sort):1, ]),
  col = consensus_palette,
  zlim = c(0, 1),
  axes = FALSE, xlab = "NMF cluster", ylab = "Samples",
  main = paste0("Consensus NMF Clustering (", k_tag, ")")
)
abline(v = mirna_cluster_boundaries[-length(mirna_cluster_boundaries)], col = "grey35", lwd = 0.8)
abline(
  h = nrow(mirna_cons_sort) - mirna_cluster_boundaries[-length(mirna_cluster_boundaries)] + 1,
  col = "grey35",
  lwd = 0.8
)
axis(1, at = mirna_cluster_midpoints, labels = paste0("C", names(mirna_cluster_sizes)), tick = FALSE)
box()
par(op)
dev.off()
```

## Save subtype assignments

```{r}
#| label: save
mirna_clusters <- predict(mirna_nmf_result)
mirna_nmf_df   <- tibble(
  sample_id = names(mirna_clusters),
  cluster   = as.integer(mirna_clusters)
)

if ("file_name" %in% names(mirna_sample_metadata)) {
  mirna_nmf_df <- mirna_nmf_df |>
    left_join(mirna_sample_metadata, by = c("sample_id" = "file_name")) |>
    relocate(
      sample_id,
      any_of(c(
        "file_id",
        "case_submitter_id",
        "case_id",
        "sample_submitter_id",
        "sample_type",
        "sample_type_code",
        "aliquot_barcode"
      )),
      cluster
    )
}

write.csv(mirna_nmf_df, file.path(mirna_data_dir, paste0("mirna_nmf_subtypes_", k_tag, ".csv")), row.names = FALSE)

nrow(mirna_nmf_df)   
```

## Sanity checks

```{r}
#| label: sanity
head(mirna_nmf_df)
table(mirna_nmf_df$cluster)          

mirna_sample_totals <- colSums(mirna_nmf_input)
png(file.path(mirna_plot_dir, paste0("mirna_counts_by_cluster_", k_tag, ".png")), width = 1000, height = 800, res = 120)
boxplot(mirna_sample_totals ~ mirna_nmf_df$cluster,
        ylab = "total miRNA counts", xlab = "cluster")
dev.off()
```

## Differential marker miRNAs

Real subtypes should have cluster-specific markers. This one-vs-rest screen ranks each miRNA by its mean expression difference in one cluster versus all other clusters, then draws heatmaps from the top up-regulated markers per cluster.

```{r}
#| label: differential-markers
#| warning: false
top_markers_per_cluster <- 5

marker_samples <- mirna_nmf_df |>
  filter(sample_id %in% colnames(mirna_lcpm)) |>
  mutate(cluster = factor(cluster, levels = sort(unique(cluster)))) |>
  arrange(cluster, sample_id)

marker_expr <- mirna_lcpm[, marker_samples$sample_id, drop = FALSE]
cluster_levels <- levels(marker_samples$cluster)

marker_stats <- purrr::map_dfr(cluster_levels, function(cl) {
  in_cluster <- marker_samples$cluster == cl
  p_values <- apply(marker_expr, 1, function(x) {
    suppressWarnings(wilcox.test(x[in_cluster], x[!in_cluster])$p.value)
  })

  tibble(
    cluster = as.integer(as.character(cl)),
    miRNA = rownames(marker_expr),
    mean_in_cluster = rowMeans(marker_expr[, in_cluster, drop = FALSE]),
    mean_other_clusters = rowMeans(marker_expr[, !in_cluster, drop = FALSE]),
    log2fc_vs_rest = mean_in_cluster - mean_other_clusters,
    p_value = unname(p_values)
  )
}) |>
  group_by(cluster) |>
  mutate(fdr = p.adjust(p_value, method = "BH")) |>
  ungroup() |>
  arrange(cluster, desc(log2fc_vs_rest))

top_markers <- marker_stats |>
  filter(log2fc_vs_rest > 0) |>
  group_by(cluster) |>
  slice_max(order_by = log2fc_vs_rest, n = top_markers_per_cluster, with_ties = FALSE) |>
  ungroup() |>
  arrange(cluster, desc(log2fc_vs_rest))

readr::write_csv(
  marker_stats,
  file.path(mirna_data_dir, paste0("mirna_differential_markers_", k_tag, ".csv"))
)
readr::write_csv(
  top_markers,
  file.path(mirna_data_dir, paste0("mirna_top_markers_", k_tag, ".csv"))
)

top_markers |>
  select(cluster, miRNA, log2fc_vs_rest, fdr)
```

```{r}
#| label: marker-heatmaps
#| fig-width: 12
#| fig-height: 7
marker_ids <- unique(top_markers$miRNA)
marker_z <- marker_expr[marker_ids, marker_samples$sample_id, drop = FALSE] |>
  t() |>
  scale() |>
  t()
marker_z[is.na(marker_z)] <- 0

marker_order <- rev(unique(top_markers$miRNA))
sample_order <- marker_samples |>
  mutate(sample_index = row_number())
sample_breaks <- sample_order |>
  count(cluster, name = "n") |>
  mutate(
    boundary = cumsum(n) + 0.5,
    midpoint = cumsum(n) - n / 2 + 0.5
  )

marker_heatmap_df <- as.data.frame(marker_z) |>
  rownames_to_column("miRNA") |>
  pivot_longer(-miRNA, names_to = "sample_id", values_to = "z") |>
  left_join(sample_order |> select(sample_id, cluster, sample_index), by = "sample_id") |>
  mutate(
    miRNA = factor(miRNA, levels = marker_order),
    z_clamped = pmax(pmin(z, 2.5), -2.5)
  )

sample_marker_heatmap <- ggplot(marker_heatmap_df, aes(sample_index, miRNA, fill = z_clamped)) +
  geom_raster() +
  geom_vline(
    data = sample_breaks |> filter(row_number() < n()),
    aes(xintercept = boundary),
    inherit.aes = FALSE,
    color = "grey45",
    linewidth = 0.25
  ) +
  scale_x_continuous(
    breaks = sample_breaks$midpoint,
    labels = paste0("C", sample_breaks$cluster),
    expand = c(0, 0)
  ) +
  scale_fill_gradient2(
    low = poster_blue,
    mid = poster_mid,
    high = poster_red,
    midpoint = 0,
    limits = c(-2.5, 2.5),
    name = "row z"
  ) +
  labs(
    title = paste0("Top miRNA markers by NMF cluster (", k_tag, ")"),
    subtitle = paste0("Top ", top_markers_per_cluster, " up-regulated miRNAs per cluster; samples ordered by cluster"),
    x = "NMF cluster",
    y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 8),
    plot.title = element_text(face = "bold")
  )

ggsave(
  file.path(mirna_plot_dir, paste0("mirna_marker_heatmap_", k_tag, ".png")),
  sample_marker_heatmap,
  width = 12,
  height = 7,
  dpi = 300
)

cluster_mean_heatmap_df <- marker_heatmap_df |>
  group_by(miRNA, cluster) |>
  summarize(mean_z = mean(z), .groups = "drop") |>
  mutate(mean_z_clamped = pmax(pmin(mean_z, 2.5), -2.5))

cluster_marker_heatmap <- ggplot(cluster_mean_heatmap_df, aes(cluster, miRNA, fill = mean_z_clamped)) +
  geom_tile(color = "white", linewidth = 0.2) +
  scale_fill_gradient2(
    low = poster_blue,
    mid = poster_mid,
    high = poster_red,
    midpoint = 0,
    limits = c(-2.5, 2.5),
    name = "mean row z"
  ) +
  labs(
    title = paste0("Cluster-average expression of top miRNA markers (", k_tag, ")"),
    x = "NMF cluster",
    y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid = element_blank(),
    axis.text.y = element_text(size = 8),
    plot.title = element_text(face = "bold")
  )

ggsave(
  file.path(mirna_plot_dir, paste0("mirna_marker_cluster_mean_heatmap_", k_tag, ".png")),
  cluster_marker_heatmap,
  width = 6,
  height = 7,
  dpi = 300
)

sample_marker_heatmap
cluster_marker_heatmap
```
